Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis

### 2015-03-03 ebayes_rnaseq_heterosis.tex
through simulation study

%\newcommand{\Stan}{{\tt Stan}}
%\newcommand{\Stan}{Stan}
\newcommand{\RStan}{{\tt RStan}}
\newcommand{\edgeR}{{\tt edgeR}}
\newcommand{\baySeq}{{\tt baySeq}}
\newcommand{\ShrinkBayes}{{\tt ShrinkBayes}}

\title{Empirical Bayes analysis of RNA-seq data for detection of gene expression heterosis}

@@

\begin{abstract}
Heterosis, or hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents. Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis of heterosis is still largely unknown. In an effort to understand phenotypic heterosis at the molecular level, researchers have begun to measure transcript abundance levels of thousands of genes in parental inbred lines and their hybrid offspring using RNA sequencing (RNA-seq) technology.  The resulting data allow researchers to search for evidence of gene expression heterosis as one potential molecular mechanism underlying heterosis of agriculturally important traits.  The null hypotheses of greatest interest in testing for gene expression heterosis are composite null hypotheses that are difficult to test with standard statistical approaches for RNA-seq analysis. To address these shortcomings, we develop a hierarchical model based on a negativeb binomial data model and draw inferences using a computationally tractable empirical Bayes approach to inference. We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment with our newly proposed methodology. This article has supplementary material online.
\end{abstract}

\begin{keywords}
Empirical Bayes; Gene expression; Heterosis; Hierarchical model; Negative binomial; RNA-seq.
\end{keywords}

@@

Recently, \cite{ji2014estimation} introduced an approach to assess gene expression heterosis using microarray data under the assumption that these data are continuous. They built a normal hierarchical model for microarray measurements of transcript abundance that allows borrowing of information across genes to estimate means and variances. They introduced an empirical Bayes framework that first estimates model hyperparameters, then estimates the posterior distribution for gene-specific parameters conditional on those hyperparameters, and finally computes heterosis probabilities based on integrals of regions under this posterior. This development was necessary due to the composite null hypotheses in tests for heterosis. These hypotheses, which many available methods do not fully accommodate, remain a challenge in the transition from continuous measurements of transcript abundance to count-based measurements that arise from RNA sequencing (RNA-seq) technology. Building on the work of \citeauthor{ji2014estimation} with the normal data model, we construct a hierarchical model based on a negative-binomial data model. We also utilize an empirical Bayes approach to obtain \emph{maximum a posterior} (MAP) estimates of the hyperparameters and the posterior distributions for the gene-specific parameters conditional on those hyperparameters.

@@

\section{Empirical Bayes identification of gene expression heterosis from RNA-seq read counts}
\label{s:method}

An RNA sequencing (RNA-seq) experiment to identify genes demonstrating expression heterosis consists of three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H). For each variety, replicate RNA samples are isolated and assessed for quality. Complementary DNA (cDNA) libraries, consisting of short cDNA fragments derived from RNA, are constructed. Then, next generation sequencing technology is used to determine the \emph{reads}, or nucleotide sequences, in the cDNA libraries. These reads are processed using bioinformatic algorithms to match the reads to genes, or specific gene transcripts, exons, microRNAs, etc. The results of read processing are summarized by a gene $\times$ sample matrix of counts. See \cite{datta2014statistical} for more details on RNA-seq experiments and data from a statistical perspective, and see \cite{paschold2012complementation} for the biological background behind the use of RNA-seq to study gene expression heterosis.

To use RNA-seq counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the parameters using an empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate empirical Bayes posterior probabilities for heterosis.

\subsection{Hierarchical model for RNA-seq counts}
\label{s:model}

@@

For a specific gene $g$, low-parent heteorisis (LPH) of gene expression occurs when average expression in the hybrid is less than the average expression of either parent, i.e. $\delta_g < -|\alpha_g|$, and high-parent heterosis (HPH) occurs when average expression in the hybrid is greater than the average expression of either parent, i.e. $\delta_g > |\alpha_g|$. 
% and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{heterosis} of gene expression to be the event of LPH or HPH. The most important hypotheses are
% \begin{align}
% H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
% H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3},
% \label{e:hypotheses}
% \end{align}
% where $H_{g0}$ indicates no heterosis for gene $g$ and $H_{g1}$ indicates  heterosis for gene $g$. 
% We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g., 
% \begin{align}
% P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
% &\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right), \label{e:probs}
% \end{align}
% where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. 
We evaluate these probabilities based on empirical Bayes estimates of their posterior probabilities, e.g., 
\begin{equation}
%P(LPH_g|y) \approx 
P(LPH_g|y,\hat{\pi}_{MAP}) = P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi}_{MAP}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
\label{e:probs}
\end{equation}
where $\left(\delta_g^{(m)},\alpha_g^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. We construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities. Geneticists can use this list to prioritize future experiments to understand the molecular genetic mechanisms for heterosis.

\subsection{Implementation in {\tt Stan}}

@@

@@

Existing RNA-seq data analysis methods are not specifically designed to identify gene expression heterosis, but we adapted the approaches in \edgeR{}, {\tt baySeq}, and \ShrinkBayes{} to provide a measure of the strength of heterosis for each gene such that large values of this measure indicate support for heterosis.

The \edgeR{} approach can be used to test for differential expression between any two varieties based on the fit of a negative binomial log-linear model \citep{robinson2007moderated, robinson2010edgeR}. To construct a measure of heterosis, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$. Then, for each gene, we define a new $p$-value denoted as $p_g$ and determined as $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.  (For all relevant signifcance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no heterosis of expression for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for gene expression heterosis.

The {\tt baySeq} approach allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct. In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene. We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$. Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the only hypotheses that allow for heterosis.

The \ShrinkBayes{} package implements empirical Bayesian inference via integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}. A limitation of INLA is that it only provides approximations to marginal posteriors for parameters or linear combinations of parameters rather than an approximation to the full posterior. Thus, we cannot directly estimate posterior probabilities involving nonlinear functions of $\mu_{g1}$, $\mu_{g2}$, and $\mu_{g3}$, such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2})|y)$. To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} using the approach in \cite{ji2014estimation}. 
Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the average of parental effects, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental effect difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect relative to the average parental effect. After fitting the model with \ShrinkBayes{}, we constructed a measure of heterosis for each gene. For gene $g$, we used the estimated posterior means $\hat{\delta}_g$ and $\hat{\alpha}_g$ of $\delta_g$ and $\alpha_g$, respectively, along with appropriate posterior probabilities. Specifically, the measure of heterosis for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g + \alpha_g > 0 |y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.

@@

For each of the methods above, as well as the method described in Section \ref{s:gene_heterosis}, we sorted genes according to the computed measure of the strength of evidence for heterosis. From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with heterosis, as defined in Section \ref{s:sim_data}  \citep{landau2013dispersion}, 
% From Dan, "why this citation?"
from those without heterosis. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
\begin{figure}[htbp]
\centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
\caption{Representative ROC curves for false positive rates below 0.1 for the approaches using \edgeR{}, {\tt baySeq},  and \ShrinkBayes{} described in Section \ref{s:alternative}, along with the independence and dependence methods using \RStan{} described in Section \ref{s:gene_heterosis}.}
\label{f:roc}
\end{figure}
For this simulation, we can see that the approaches based on the model in Section \ref{s:model} $-$ i.e. both \RStan{} approaches and the \ShrinkBayes{} approach $-$ provide the best true positive rate for a given false positive rate. Also, as expected, as the sample size increases, our ability to distinguish genes with heterosis from genes without improves.

Figure \ref{f:auc} provides the area under the ROC curve below a false positive rate of 0.1 across the 10 simulations for each of the 3 different sample sizes. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different sample sizes for the approaches using \edgeR{}, {\tt baySeq},  and \ShrinkBayes{} described in Section \ref{s:alternative} , along with the independence and dependence methods using \RStan{} described in Section \ref{s:gene_heterosis}. Each line is a different simulation while the blue box indicates mean AUC and mean AUC plus and minus one standard error across the simulations.}
\label{f:auc}
\end{figure}
The pattern appears to hold, with the \RStan{} and \ShrinkBayes{} techniques outperforming the other two. While there does not appear to be a significant difference between \ShrinkBayes{} and the independence model using \RStan{}, performance for \RStan{} appears to decrease for the dependence model where an unstructured covariance is estimated, at least for smaller sample sizes.

@@

Geneticists speculate that gene expression heterosis is one possible explanation of phenotypic heterosis of traits, such as plant height or grain yield. Because of the unwieldy composite null hypothesis in \eqref{e:hypotheses}, existing methods for identifying gene expression heterosis based on RNA-seq data are not directly applicable for detecting heterosis genes. \cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data. We followed their approach, modified to allow for RNA-seq read counts as measures of transcript abundance. We developed an empirical Bayes approach based on obtaining MAP estimates for hyperparameters followed by MCMC to estimate gene-specific parameters. This methodology can be conveniently implemented in the statistical software, {\tt Stan}. The empirical Bayes posteriors can be used to estimate posterior probabilities of high and low parent heterosis.

>>>>

%\newcommand{\Stan}{{\tt Stan}}
\newcommand{\Stan}{Stan}
\newcommand{\RStan}{{\tt RStan}}
\newcommand{\edgeR}{{\tt edgeR}}
\newcommand{\baySeq}{{\tt baySeq}}
\newcommand{\ShrinkBayes}{{\tt ShrinkBayes}}
\newcommand{\RNAseq}{RNAseq}

\title{Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis}

@@

\begin{abstract}
Heterosis, or hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents. Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis of heterosis is still largely unknown. In an effort to understand phenotypic heterosis at the molecular level, researchers have begun to measure transcript abundance levels of thousands of genes in parental inbred lines and their hybrid offspring using RNA sequencing (\RNAseq{}) technology.  The resulting data allow researchers to search for evidence of gene expression heterosis as one potential molecular mechanism underlying heterosis of agriculturally important traits.  The null hypotheses of greatest interest in testing for gene expression heterosis are composite null hypotheses that are difficult to test with standard statistical approaches for \RNAseq{} analysis. To address these shortcomings, we develop a hierarchical model based on a negativeb binomial data model and draw inferences using a computationally tractable empirical Bayes approach to inference. We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment with our newly proposed methodology. This article has supplementary material online.
\end{abstract}

\begin{keywords}
Empirical Bayes; Gene expression; Heterosis; Hierarchical model; Negative binomial; \RNAseq{}.
\end{keywords}

@@

Recently, \cite{ji2014estimation} introduced an approach to assess gene expression heterosis using microarray data under the assumption that these data are continuous. They built a normal hierarchical model for microarray measurements of transcript abundance that allows borrowing of information across genes to estimate means and variances. They introduced an empirical Bayes framework that first estimates model hyperparameters, then estimates the posterior distribution for gene-specific parameters conditional on those hyperparameters, and finally computes heterosis probabilities based on integrals of regions under this posterior. This development was necessary due to the composite null hypotheses in tests for heterosis. These hypotheses, which many available methods do not fully accommodate, remain a challenge in the transition from continuous measurements of transcript abundance to count-based measurements that arise from RNA sequencing (\RNAseq{}) technology. Building on the work of \citeauthor{ji2014estimation} with the normal data model, we construct a hierarchical model based on a negative-binomial data model. We also utilize an empirical Bayes approach to obtain \emph{maximum a posterior} (MAP) estimates of the hyperparameters and the posterior distributions for the gene-specific parameters conditional on those hyperparameters.

@@

\section{Empirical Bayes identification of gene expression heterosis from \RNAseq{} read counts}
\label{s:method}

An RNA sequencing (\RNAseq{}) experiment to identify genes demonstrating expression heterosis consists of three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H). For each variety, replicate RNA samples are isolated and assessed for quality. Complementary DNA (cDNA) libraries, consisting of short cDNA fragments derived from RNA, are constructed. Then, next generation sequencing technology is used to determine the \emph{reads}, or nucleotide sequences, in the cDNA libraries. These reads are processed using bioinformatic algorithms to match the reads to genes, or specific gene transcripts, exons, microRNAs, etc. The results of read processing are summarized by a gene $\times$ sample matrix of counts. See \cite{datta2014statistical} for more details on \RNAseq{} experiments and data from a statistical perspective, and see \cite{paschold2012complementation} for the biological background behind the use of \RNAseq{} to study gene expression heterosis.

To use \RNAseq{} counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the parameters using an empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate empirical Bayes posterior probabilities for heterosis.

\subsection{Hierarchical model for \RNAseq{} counts}
\label{s:model}

@@

For a specific gene $g$, low-parent heteorisis (LPH) of gene expression occurs when average expression in the hybrid is less than the average expression of either parent, i.e. $\delta_g < -|\alpha_g|$, and high-parent heterosis (HPH) occurs when average expression in the hybrid is greater than the average expression of either parent, i.e. $\delta_g > |\alpha_g|$. 
% and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{heterosis} of gene expression to be the event of LPH or HPH. The most important hypotheses are
% \begin{align}
% H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
% H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3},
% \label{e:hypotheses}
% \end{align}
% where $H_{g0}$ indicates no heterosis for gene $g$ and $H_{g1}$ indicates  heterosis for gene $g$. 
% We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g., 
% \begin{align}
% P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
% &\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right), \label{e:probs}
% \end{align}
% where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. 
We evaluate these probabilities based on empirical Bayes estimates of their posterior probabilities, e.g., 
\begin{equation}
%P(LPH_g|y) \approx 
P(LPH_g|y,\hat{\pi}_{MAP}) = P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi}_{MAP}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
\label{e:probs}
\end{equation}
where $\left(\delta_g^{(m)},\alpha_g^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. HPH probability is defined similarly with the inequality reversed and without the negative sign. We construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities. Geneticists can use this list to prioritize future experiments to understand the molecular genetic mechanisms for heterosis.

\subsection{Implementation in \Stan{}}

@@

\subsection{Implementation in \ShrinkBayes{}}

\ShrinkBayes{} utilizes integrated nested Laplace approximation (INLA) \citep{rue2009approximate} in combination with empirical Bayes ideas \citep{van2014shrinkbayes}. One limitation with inferential methods based on INLA is that all distributions, except for the data distribution, must have tails as light or lighter than the normal density. Thus, we cannot implement the Laplace priors for the half-parental difference ($\alpha_g$) and the hybrid difference ($\delta_g$) and instead use normal priors in this situation. An additional limitation is that INLA provides approximations to marginal posteriors for parameters or linear combinations of parameters, but not an approximation to the full posterior. Since we are interested in non-linear quantities such as $P(\delta_g > |\alpha_g||y)$, we cannot compute these directly using \ShrinkBayes{}. Instead, for \ShrinkBayes{}, we calculate heterosis probabilities conditional on posterior means for $\hat{\delta}_g$ and $\hat{\alpha}_g$, e.g. 
\[ 
% P_{\ShrinkBayes{}}(LPH_g|y) \equiv \left\{ \begin{array}{ll}
% P(\delta_g>-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
% P(\delta_g>\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
% 0 & \mbox{otherwise}
% \end{array} \right.
P_{\ShrinkBayes{}}(LPH_g|y) \equiv P(\delta_g>-\alpha_g|y)\mathrm{I}\left(\hat{\delta}_g > -\hat{\alpha}_g \ge 0\right) + P(\delta_g>\alpha_g|y) \mathrm{I}\left(\hat{\delta}_g > \hat{\alpha}_g \ge 0\right).
%P_{\ShrinkBayes{}}(HPH|y) \equiv \left\{ \begin{array}{ll}
%P(\delta_g<-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
%(\delta_g<\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
%0 & \mbox{otherwise}
%\end{array} \right.
\]
HPH probability is defined similarly with all inequalities reversed. As before, we construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities.
%the heterosis probability for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g>\alpha_g|y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.

@@

Existing \RNAseq{} data analysis methods are not specifically designed to identify gene expression heterosis, but we adapted the approaches in \edgeR{} and {\tt baySeq} to provide a measure of the strength of heterosis for each gene such that large values of this measure indicate support for heterosis.

\edgeR{} can be used to test for differential expression between any two varieties based on the fit of a negative binomial log-linear model \citep{robinson2007moderated, robinson2010edgeR}. To construct a measure of heterosis, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$. Then, for each gene, we define a new $p$-value denoted as $p_g$ and determined as $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.  (For all relevant signifcance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no heterosis of expression for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for gene expression heterosis.

{\tt baySeq} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct. In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene. We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$. Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the only hypotheses that allow for heterosis.

@@

For each of the methods above, as well as the method described in Section \ref{s:gene_heterosis}, we sorted genes according to the computed measure of the strength of evidence for heterosis. From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with heterosis, as defined in Section \ref{s:sim_data}, 
% From Dan, "why this citation?"
from those without heterosis \citep{landau2013dispersion}. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
\begin{figure}[htbp]
\centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
\caption{Representative ROC curves for false positive rates below 0.1 for the approaches using \edgeR{}, {\tt baySeq},  and \ShrinkBayes{} described in Section \ref{s:alternative}, along with the independence and dependence methods using \RStan{} described in Section \ref{s:gene_heterosis}.}
\label{f:roc}
\end{figure}
For this simulation, we can see that the approaches based on the model in Section \ref{s:model}, i.e. \RStan{} and \ShrinkBayes{}, provide the best true positive rate for a given false positive rate. Also, as expected, as the sample size increases, our ability to distinguish genes with heterosis from genes without improves.

Figure \ref{f:auc} provides the area under the ROC curve below a false positive rate of 0.1 across the 10 simulations for each of the 3 different sample sizes. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different sample sizes for the approaches using \edgeR{}, {\tt baySeq},  and \ShrinkBayes{} described in Section \ref{s:alternative} , along with the independence and dependence methods using \RStan{} described in Section \ref{s:gene_heterosis}. Each line is a different simulation while the blue box indicates mean AUC and mean AUC plus and minus one standard error across the simulations.}
\label{f:auc}
\end{figure}
The pattern appears to hold, with the \RStan{} and \ShrinkBayes{} techniques outperforming the other two. %While there does not appear to be a significant difference between \ShrinkBayes{} and the independence model using \RStan{}, performance for \RStan{} appears to decrease for the dependence model where an unstructured covariance is estimated, at least for smaller sample sizes.

@@

Geneticists speculate that gene expression heterosis is one possible explanation of phenotypic heterosis of traits, such as plant height or grain yield. Because of the unwieldy composite null hypothesis in \eqref{e:hypotheses}, existing methods for identifying gene expression heterosis based on \RNAseq{} data are not directly applicable for detecting heterosis genes. \cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data. We followed their approach, modified to allow for \RNAseq{} read counts as measures of transcript abundance. We developed an empirical Bayes approach based on obtaining MAP estimates for hyperparameters followed by MCMC to estimate gene-specific parameters. This methodology can be conveniently implemented in the statistical software, {\tt Stan}. The empirical Bayes posteriors can be used to estimate posterior probabilities of high and low parent heterosis.
